A Critical Review of Sophistication and Algorithmic Model Selection

### 2014-12-19 Facticity/Facticity/paper.tex
...

\begin{abstract}
\sdr{Emphasize ongoing attempts to formalize meaningful information, ``recurring dream''. Also in the introduction} \sdr{Don't single out sophistication}
Kolmogorov complexity is a sound expression of the information content of a binary string, but it leaves open the issue of \emph{meaningful information}. The strings with maximal information content are not the ones we would regard as interesting, or meaningful. One attempt to capture the idea of meaningful information is \emph{sophistication}. It is a simple idea with many attempted formalizations: describe the string efficiently with a two-part coding approach, and take the complexity of the model as a measure of the amount of meaningful information. We describe three requirements that any formalization must meet in order to satisfy the expectations that come with the idea of sophistication: invariance, unboundedness and efficiency. We show that no currently published proposal meets these requirements \pb{TODO}, and conjecture that no formalization of the idea of sophistication can hope to do so. We conclude by reviewing the consequences of this conjecture.
\end{abstract}

\sdr{Get conjecture right}

@@

Imagine an old-fashioned, analog television set which can be tuned to three frequencies. The first produces a blank screen, the second an ordinary TV show, and the third white noise. Judging by their Kolmogorov complexity, these  broadcasts have increasing information content. The first is easiest to describe, the second is harder but still contains a great deal of regularity, and the third cannot be compressed. However, intuitively speaking the second signal clearly conveys more meaningful information than the other two. This notion has proved far harder to capture in formal terms than the simple notion of `raw' information content. \sdr{Emphasize recurring dream here}

One of the most frequently explored ideas for capturing the notion of meaningful information is \emph{sophistication}. The idea behind sophistication is to select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and to take the complexity of only the model part as a measure of the meaningful information. The assumption is that both simple and random data have small models: \sdr{the intuition is that small models suffice to describe both simple and random data, but not the meaningful information we are attempting to quantify.} in the case of the simple data, a small model describes all regularity, and no further information is required, and in the case of the random data, the data itself is its shortest representation, and the model is a `print function' which simply reproduces its input. If the dataset is more interesting, however, containing both regularity and randomness, its best representation will have a more complex model. \sdr{slightly repetitive} A key assumption of the idea of sophistication is that we can perform objective model selection for a given string. That is: there is a function that, given a string, selects a single unique model that can be used to compress this string optimally. Moreover, any elements of this function that are chosen ad hoc can be changed without significantly affecting the size of the preferred model.

In the case of the Kolmogorov complexity, this holds true; we can change the universal Turing machine used in its definition and the complexities returned will not change by more than a constant. This is known as \emph{invariance}. Because Kolmogorov complexity is invariant, we can view it as an intrinsic property of the data, rather than simply an arbitrary function computed on it. If a formalization of sophistication is ever to be held in the same regard as the Kolmogorov complexity, it must be subject to a similar principle of invariance. \sdr{I feel it's not about regard}

The unifying principle over all the notions we review in this paper is simple. We start with an object expressed as a finite binary string. We represent this object, hoping to compress it, in two parts, the first is a model, and the second an `input' to this model. How the two parts are encoded, and what the class of models is are details that differ between variants. We know, however, that for successful representations, those that capture all structure in the string, the length of the two-part code should approach the Kolmogorov complexity. We take the set of all \emph{sufficient statistics}---the models of those representations that are `close enough' to the Kolmogorov complexity---and call the one with the lowest complexity the \emph{minimal sufficient statistic}. The complexity of the minimal sufficient statistic is the sophistication. \sdr{But I thought in some of our settings we use a different baseline than KC!}

The intuition is that both very simple and very random data have small, simple programs. The strings with much structure and many interesting properties are those for which the best model is large.\sdr{Repetitive} Two assumptions are crucial to this view: \sdr{This bit should be more solid}
\begin{itemize}
  \item The two-part coding efficiently \sdr{what does efficiently mean?} separates the structure \sdr{undefined} and the noise into distinct parts. In the optimal representation the model contains only `structural information' and the noise contains only the `uninteresting' information.
  \item This principle is invariant. If we change, for instance, the universal Turing machine used to unpack our descriptions, the minimal sufficient statistic will not change significantly.
\end{itemize}

In this article we will show that most proposals do not achieve these goals. For one remaining case, we can offer no such proof, but there exists no proof to the contrary either. \sdr{Say this more strongly: we believe that ``fixes'' of the problems with the approach are obfuscation rather than real solutions}  We offer several insights that suggest \sdr{insights that suggest stuff sounds very weak} that the idea of an objective sophistication function is an unattainable ideal.

@@

We will first briefly discuss the different variants of sophistication in a historical context, to give the reader a simple, chronological overview, without discussing too much technical detail. \sdr{We first provide a chronological overview of the variants of sophistication and discuss their relationships.} In the main body of the paper, we will define each variant rigorously, and in non-chronological order.

@@

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity \sdr{why?}. One interesting difference is that Koppel takes the description of the data under a variant of Kolmogorov complexity, which consists of a description of a model and that model's input, and takes the size of the model part as the sophistication. In contrast, Kolmogorov's minimal sufficient statistic is built on a second, two-part, representation in addition to the Kolmogorov complexity's one-part representation. This difference in perspective is subtle but relevant to the discussion ahead.\sdr{Good point but currently much too hard to understand}

\sdr{Following bit is currently pretty vague. I think the collapse idea deserves a better introduction} While Koppel expands the model class from finite sets to functions, he maintains a restriction that the functions be total. This is a recurring theme: if the model class is expanded to all total functions, the the universal Turing machine itself becomes a model, so that we always have a two-part description with a small model, to within a constant of the Kolmogorov complexity. In most treatment this `collapse' to the universal model is avoided by excluding it from the model class, using some restriction. This acts as a kind of force, pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes a balancing act, trying to match two forces which by themselves push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{todo}. Working from the perspective of statistical mechanics, they arrived at a notion remarkably similar to Sophistication. Effective complexity describes abinary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) \sdr{I really think the ensemble is just the set} and the (Shannon) entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{}, Gell-Mann and Lloyd select the decisive two-part representation from those achieving minimal code-length based on ``the average quantities judged to be important'', which would seem to make the value highly subjective and context-dependent.

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character \sdr{Source?}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model. He balances this force, by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.\sdr{This balancing act is not yet clear to the reader}

@@

Given the philosophical underpinnings of sophistication, these three requirements are the very least that we could expect. We will show that no current proposal satisfies all three. In the is no proposal for which invariance has been proved to be satisfied. This leads us to the following conjecture:

\begin{conj}
Let $\s_g$ be a sophistication function constructed as in Definition~\ref{definition:sophistication}, built on some acceptable numbering $g$, and some fixed model class $\M$ and threshold function $\tau$ (left out of the sub/superscript for the sake of clarity). \pb{dit kan mooier}

There is no combination of $\M$ and $\tau$ such that the following all hold: 
\begin{enumerate}
  \item $\forall x : \left|\s_g(x) - \s_h(x)\right| \leq c$ for some constant $c$. (invariance)
  \item $\forall c \exists x : \s_g(x) \geq c$ and $\forall c \exists x : L^\M(x) \s_g(x) \geq c$. (unboundedness)
  \item $forall x : L^\M(x) = K(x)$ if the representations in $\M$ are self-delimiting, or $\forall x : L^\M(x) = C(x)$ otherwise. (efficiency)
\end{enumerate}
\end{conj}
In more informal terms, we conjecture that no approach following the basic principle of sophistication---that is, model selection followed by measuring of the model complexity---can satisfy all three requirements.

@@

\begin{itemize}
\item Model class is $\T$
\item Threshold function $\tau(x)=L^\M(x)+c$
\item Description method for models is required to be prefix-free, but
  can otherwise be chosen arbitrarily.
\item Koppel uses monotone complexity, and monotone functions and discusses 
  infinite strings, TODO figure out if and how this affects our results.
\end{itemize}

@@

\begin{itemize}
\item Model class is \emph{ensembles}. Equivalent to $\K$.
\item Description method for models is $L(m) = K(m)$ (the minimization is not mentioned explicitly, but we'll give them the benefit of the doubt).
\item The threshold function is equal to the minimum two-part code: $\tau(x)=L^m(x)$.
\item Gell-mann and Lloyd acknowledge that representations exist with small and large models. The choice from the candidate set is made by applying a time-bound to the computation from the string to the model, but not to the decoding of the model itself. The idea is that this forces `deep' information into the model, and shallow information into the data. Presumably, the representation with the smallest model code under this constraint is chosen.   
\end{itemize} 
\pb{We can still manipulate the UTM so that all non-singleton representations are pushed out of the candidate set. Since the model side is allowed unbounded decoding time, this invalidates effective complexity.}

>>>>

\begin{abstract}
Kolmogorov complexity is a sound expression of the information content of a binary string, but it clashes with our intuitive notions of information: the strings with maximal information content are not the ones we would regard as interesting, or meaningful. One attempt to capture this idea of `interesting' information is \emph{sophistication}. It is a simple idea with many attempted formalizations: describe the string efficiently with a two-part coding approach, and take the complexity of the model as a measure of the amount of interesting information. Kolmogorov complexity is invariant to the ad-hoc choices in its construction: change them, and the result stays the same, up to a constant. We review the history of sophistication critically, asking whether invariance exists for the proposed sophistication functions as well. In most cases, we can show that it doesn't. We conjecture that no formalization of sophistication can be invariant, and we briefly review the consequences of this conjecture.
\end{abstract}

@@

Imagine an old-fashioned, analog television set which can be tuned to three frequencies. The first produces a blank screen, the second an ordinary TV show, and the third white noise. Judging by their Kolmogorov complexity, these  broadcasts have increasing information content. The first is easiest to describe, the second is harder but still contains a great deal of regularity, and the third cannot be compressed. However, intuitively speaking the second signal clearly conveys more meaningful information than the other two. This notion has proved far harder to capture in formal terms than the simple notion of `raw' information content.

Over the years one idea in particular has been proposed in many different forms. We select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and we take the complexity of only the model part as a measure of the meaningful information. If we can find a single two-part description that we can objectively call the optimal description of the data, the complexity of the data part gives us an objective property of the data. While this idea has been proposed many times in many different guises, we will, for the purposes of this paper refer to it as \emph{sophistication}\footnotemark. The general problem of how to characterize the `richness' of the information in a string rather than the plain information content (to which sophistication is but one proposed solution), we will call the `problem of interesting information'.

\footnotetext{As the historical overview shows, sophistication is only one name for this idea, and not the first. The alternatives, however, are more awkward, and it has been used by different authors for different realizations of the same principle.}

The intuitive idea of sophistication makes one crucial demand: we can find an objectively `true' two-part description. In the framework of Komogorov complexity two observers cannot disagree by more than a constant about the information content of a dataset. In formal terms: the Kolmogorov complexity of a dataset $x$ is defined as the shortest effective description for $x$ in some Turing complete description language. This description language is usually expressed as a universal Turing machine $U(\bar\imath y) = \phi_i(y)$ where $\phi_i$ is the $i$'th partial computable function in some effective enumeration, and the $\bar\imath$ is a self-delimiting encoding of $i$. If two people disagree about the shortest effective description of $x$, we can interpret this as them using different description languages, ie. different universal Turing machines. The crucial aspect of Kolmogorov complexity, the starting theorem that makes it interesting as a function at all, is \emph{invariance}: two people using different description languages will get the same value for a string's Kolmogorov complexity, albeit up to a fixed constant. Without invariance, Kolmogorov complexity would be an arbitrary function. With invariance, we can be sure that we are talking about an intrinsic property of the data. The same consideration must be given to each proposal for formalising sophistication: is it invariant?

There are many approaches to sophistication, but the following pattern is common to all of them. We start with a space of \emph{models}. There can be finite sets, computable probability distributions or simple functions from strings to strings. The function perspective is the most generic, all other model classes can be translated to it. The approach to sophistication is to find a representation for a dataset $x$ in two parts: a computable function $f$ and and input $y$, such that $f(y) = x$. As with Kolmogorov complexity, the shorter a description is, the better. We take the shortest description and those close to it, and consider them \emph{candidates}. How close a description must be to the optimum differs between variants, it is often a pre-determined constant value. We then choose a single description from the candidate set and return its \emph{model complexity} as the sophistication. The method for determining model complexity varies. In almost all variants the candidate with the smallest model complexity determines the sophistication.

The implicit assumption behind sophistication is that this approach will separate the description of the string into a sturctural part, representing its regularities, and a stochastic part representing its non-deterministic properties. In other words, the string is cast as a random choice from a model for which it is `typical'.

In this article we will show that most proposals are not invariant. For one remaining case, we can offer no such proof, but no proof of invariance exists either. This is likely due to the complex nature of the model class chosen, and the difficulty of resolving the matter of invariance shows how complex such a sophistication measure would be to build on, even if it did prove to be invariant.

@@

We first provide a chronological overview of the variants of sophistication. In the body of the paper, we will define each variant rigorously, and in non-chronological order.

@@

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity. This is likely to extend the theory to infinte strings, an approach which has been abandoned by subsequent authors.

While Koppel expands the model class from finite sets to functions, he maintains a restriction that the functions be total. This is a recurring theme: if the model class is expanded to all computable functions, the the universal Turing machine $U$ itself becomes a model. If we assume that $U$ computes the function $\phi_u$ in the standard enumeration of computable functions, and that $\bar\imath y$ is a shortest description for $x$ on $U$, then $\br{u}\br{\imath}y$ is also a description for $x$. This description has a constant model size ($|\br{u}|$), ie. every $x$ has constant sophistication. This `collapse' to a universal modle is the reason that most treatments of sophistication constrain the model class to exclude the universal Turing machine. This acts as a kind of force, pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes a balancing act, trying to match two forces which by themselves would push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that firts the mold of sophistication. Effective complexity describes a binary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) and the entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{gell2004nonextensive}, Gell-Mann and Lloyd select the decisive two-part representation from the candidates based on ``the average quantities judged to be important'', which seems to be a concious choice to make the measure subjective and context-dependant (sacrificing invariance).

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character. \cite{kolmogorov1983combinatorial}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model. He balances this force by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.\sdr{This balancing act is not yet clear to the reader}

@@

Given the philosophical underpinnings of sophistication, these three requirements are the very least that we could expect. We will show that no current proposal satisfies all three.

For all variants but one, we have shown that all measures that are invariant are bounded. This leads us to the following conjecture. Let $\gamma(n)$ be a function representing the extent to which we care about differences between functions. That is, we will call two functions $f$ and $g$ $\gamma$-equal if $|f(x) - g(x)| \in O(\gamma(f(x)))$.

\begin{conj}
Let $\s_g$ be a sophistication function constructed as in Definition~\ref{definition:sophistication}, built on some acceptable numbering $g$, and some fixed model class $\M$ and threshold function $\tau$ (omitted from the sub/superscript for the sake of clarity).

For any sophistication function so constructed, and any granularity $\gamma$ we have that for any two acceptable numberings $g$ and $h$, either:
\begin{itemize}
  \item there exist $s_g$ and $s_h$ which are not are $\gamma$-equal (ie $s$ is not invariant),
  \item $s_g$ is $\gamma$-equal to $0$ or to the Kolmogorov complexity.
\end{itemize}
\end{conj}
In these terms, all instantiations of the Kolmogorov complexity, using different Universal Turing machines are $1$-equal. It may be unreasonable to require the sophistication to attain the same level of invariance (a log-invariant sophistication could still be useful). This conjecture allows such relaxations: if it can be shown to be true, sophistication must be considered fundamentally flawed. We hope that this conjecture will prompt authors to address the matter of invariance ciritcally, and to consider the possibility that sophistication and its ilk are an untenable proposition.

@@

\begin{itemize}
\item Model class is $\T$
\item Threshold function $\tau(x)=L^\M(x)+c$
\item Description method for models is required to be prefix-free, but
  can otherwise be chosen arbitrarily.
\item Koppel uses monotone complexity, and monotone functions and discusses 
  infinite strings, TODO figure out if and how this affects our results.
  \item Koppel has a `machine-independence result', but not one that does what we want.
\end{itemize}

@@

\begin{itemize}
\item Model class is \emph{ensembles}. Equivalent to $\K$. Subsequent literature suggests that ensembles have finite support. 
\item Description method for models is $L(m) = K(m)$ (the minimization is not mentioned explicitly, but we'll give them the benefit of the doubt).
\item The threshold function is equal to the minimum two-part code: $\tau(x)=L^m(x)$.
\item Gell-mann and Lloyd acknowledge that representations exist with small and large models. The choice from the candidate set is made by applying a time-bound to the computation from the string to the model, but not to the decoding of the model itself. The idea is that this forces `deep' information into the model, and shallow information into the data. Presumably, the representation with the smallest model code under this constraint is chosen.   
\end{itemize} 
\pb{We can still manipulate the UTM so that all non-singleton representations are pushed out of the candidate set. Since the model side is allowed unbounded decoding time, this invalidates effective complexity.}
